Mendelian randomization for causal inference in genome-wide association studies

### 2022-02-28 mr-review.Rmd
fixed section title
minor comment
minor edits
tweaks

---
title: "Mendelian randomization for causal inference in Genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

They are termed 
"two-sample" MR studies because the exposure variable and the outcome variable are measured on 
two non-overlapping cohorts of subjects [@pierce_efficient_2013]. Importantly, the SNP instruments are measured in both sets of subjects.

@@

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its use in the GWAS setting.

@@

Note that $k$ indexes SNP instruments and ranges from 1 to $K$ in Equation \@ref(eq:smm1).

@@

In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity Q test. In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate Q in Equation \@ref(eq:q).

@@

While the Q test tends to be conservative in small sample sizes, its power increases with increasing 
sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].

@bowden_mendelian_2015 adapted Egger regression, originally developed 
to detect bias in meta-analyses, to detect bias from horizontal 
pleiotropy in two-sample MR studies with GWAS summary data 
[@egger1997bias]. The slope coefficient estimates the causal effect. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].

@@

To develop MR methods that are robust to correlated horizontal pleiotropy, 
@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
accommodates both uni-directional and bi-directional effects between exposure and outcome. 
@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$. $Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which . Note that the model is agnostic with regards to which is the 
outcome and which the exposure, so we follow the authors' notation here.

@@

## Sample overlap

>>>>

---
title: "Mendelian randomization for causal inference in genome-wide association studies"
author: "Frederick J. Boehm & Xiang Zhou"
date: "`r lubridate::now()`"
bibliography: exported.bib
csl: elsevier-with-titles.csl
link-citations: yes
linkcolor: blue
output: 
  - bookdown::pdf_document2
  - bookdown::html_document2
  - bookdown::word_document2
---

@@

@@

In Equation \@ref(eq:wald), $\hat \beta_{YZ}$ and $\hat \beta_{XZ}$ are the slope 
estimates from the regressions of the outcome and exposure, respectively, on the 
SNP instrument. Since the Wald method requires only the regression coefficients, it can be used with summary data. However, the Wald method doesn't accommodate multiple SNP instruments, which limits its direct use in the GWAS setting.

@@

Note that $k$ indexes SNP instruments and ranges from $1$ to $K$ in Equation \@ref(eq:smm1).

@@

In efforts to detect the presence of horizontal pleiotropy in two-sample MR studies with GWAS summary data, @greco_m_detecting_2015 developed the between-instrument heterogeneity $Q$ test. In a meta-analysis of MR Wald ratio estimates, one per SNP instrument, they calculate $Q$ in Equation \@ref(eq:q).

@@

While the $Q$ test tends to be conservative in small sample sizes, its power increases with increasing 
sample size and increasing degree of pleiotropy [@greco_m_detecting_2015].

@bowden_mendelian_2015 adapted Egger regression, originally developed 
to detect bias in meta-analyses, to detect bias from horizontal 
pleiotropy in two-sample MR studies with GWAS summary data 
[@egger1997bias]. The slope coefficient estimates the causal effect of the exposure on the outcome. 
Under the assumption that the SNP-exposure relationship is independent 
of the SNP's pleiotropic effects, Egger's test offers a valid test of 
the null causal hypothesis and consistently estimates the causal 
effect, even if all SNPs are invalid instruments [@bowden_mendelian_2015].

@@

To develop MR methods that are robust to correlated horizontal pleiotropy, 
@brown_welch-weighted_2021 present Welch-weighted Egger regression. Their flexible model 
accommodates both uni-directional and bi-directional effects between exposure and outcome. 
@brown_welch-weighted_2021 specifies their model via Equations \@ref(eq:wwer). They denote 
the two traits by $A$ and $B$, the SNP genotypes matrix by $X$, and the unobserved confounder by $U$. $Z_U$, $Z_A$ and $Z_B$ denote the indicator vectors of zeros and ones to indicate which . Note that the model is agnostic with regards to which is the 
outcome and which the exposure, so we follow the authors' notation here. The symbol $\odot$ denotes the Hadamard elementwise product.

@@

## Sample overlap in biobanks
